Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-04 appendix.tex
restructing, moving iso to appendix

\section{Applicability of the vertically global shearing box}\label{global_corr}
In \S\ref{analytical} we used the vertically global shearing box
\citep[VGSB,][]{mcnally14} to make analytic progress. % Here, we discuss limitations of this approach.   
The main difference between the linear problem in the VGSB 
and the radially-local approximation
(Eq. \ref{lin_mass}---\ref{lin_energy}) adopted in our numerical study
is the absence of  terms associated with the radial disk structure in
the former framework. These terms appear in the energy, mass and radial momentum
equations and are reproduced here for convenience,
% Using the notation and disk models in the main text, the
% energy, density and $r$ momentum equations become 
\begin{align}
  \ii \sigma W  &=c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{\gamma}{\Gamma} \frac{d\delta
    v_z}{dz}\right) + \left[\ii k_xc_s^2(r_0,z)
  \frac{\gamma}{\Gamma} + \hat{g}_c\left.\frac{1}{\rho}\frac{\p P}{\p
      r}\right|_0\right]\delta v_x  +
\frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right),\label{gcorr_terms1}\\
\ii \sigma Q &= c_s^2(r_0,z)\left(\ii k_r + \hat{g}_c
  \left.\frac{\p\ln{\rho}}{\p r}\right|_0\right)\delta v_x + c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{d\delta
    v_z}{dz}\right),\label{gcorr_terms2} \\
\ii \sigma \delta v_x & = \ii k_x W  -
\hat{g}_c\frac{1}{c_s^2(r_0,z)}\left.\frac{1}{\rho}\frac{\p P}{\p
  r}\right|_0Q - 2\Omega(r_0,z)\delta v_y,\label{gcorr_terms3}
\end{align}
where subscript $0$ denotes evaluation at $r=r_0$.

@@

%  To obtain the
% linear problem in the VGSB, we take the radially-local 
% approximation further and formally set $\hat{g}_c=0$, $r\to x$ and
% $k_r\to k_x$. It is clear that this approximation improves with
% increasing $|k_x|$.

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
An important caveat of the VGSB is that it cannot be used in the
adiabatic limit. We explain this by setting  $\beta\to\infty$ in
Eq. \ref{vertiso_gov_nondim}, which can then be written as 
% For adiabatic perturbations ($\gamma>1$, $\beta\to\infty$) we may
% simply use the Soilberg-Hoiland critera to assess stability
% (\S\ref{solberg}). Nevertheless, we briefly consider this case in the
% local framework to demonstrate its limitations. For $\beta\to\infty$
% the parameter $\chi = 1/\gamma$ and Eq. \ref{vertiso_gov_nondim} can
% be written as   
\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
$\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
expected, a sub-adiabatically stratified disk is stable in the absence
of vertical shear.

@@

% To 
% avoid this, we must restrict our attention to short thermal relaxation
% timescales ($\beta\lesssim1$) and/or neutrally-stratified disks
% ($\gamma=\Gamma$) when $q\neq0$.

\subsection{Effect of global radial gradients}
% We may also reverse the above procedure by reinstating the `global
% correction' terms proportional to $\hat{g}_c$ in the linear VGSB
% equations. Of course, doing so implies the basic state is not uniform
% in $x$, which would be inconsistent with the full VGSB equations in
% equilibrium. Nevertheless, we illustrate in Fig. \ref{gcorr_compare} the effect of
% these global radial gradients on the growth rate of the fundamental
% VSI.

% The global correction terms do not change the qualitative dependence
% of the fundamental VSI growth rates on the thermal relaxation
% timescale. The difference is negligble for $\beta\lesssim 1$ (i.e.
% fast thermal relaxation). The critical relaxation timescale for the
% fundamental VSI is unaffected by these global radial radients.

% However, in the adiabatic limit $\beta\to\infty$ the linear VGSB
% equations give a positive growth rate as discussed in
% \S\ref{analytic_adia}, which is inconsistent with the second
% Solberg-Hoiland criterion (Eq. \ref{solberg2}). The correct behaviour
% ($\nu\to0$ as $\beta\to\infty$) is obtained with global correction.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/gcorr_compare} 
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation time $\beta$. The `Analytical (VGSB)' curve
    is calculated from the dispersion relation Eq. \ref{relax_disp};
    the `Numerical (VGSB)' curve is obtained by numerically solving
    Eq. \ref{lin_mass}---\ref{lin_energy} with $\hat{g}_c=0$
    (which defines the linear VGSB equations); and the `Numerical
    (Full)' curve is obtained by numerically solving
    Eq. \ref{lin_mass}---\ref{lin_energy}  with $\hat{g}_c=1$, which
    accounts for the radial disk structure.  
% computed from the linear VGSB equations used in
%     the main text (dotted) and that including additional global
%     radial gradient terms in
%     Eq. \ref{gcorr_terms1}---\ref{gcorr_terms3} (solid). 
    The disk parameters are $\Gamma=1.011$,
    $\gamma=1.4$ and $(p,q,\epsilon)=(-1.5,-1,0.05)$. The perturbation
    wavenumber is $\khat=30$. 
    \label{gcorr_compare}}  
\end{figure}

%nevertheless we have calculated
%plot growth rate v.s. bcool

\section{Full governing equation for vertically isothermal disks in
  the VGSB}\label{adia_improve}
% \section{Improving the nearly-Keplerian approximation for adiabatic
%   disks}\label{adia_improve}
In \S\ref{approx_gov} we made the replacement $D\to\Omega_k^2$ before
eliminating variables to obtain a single equation for $\delta v_z$,
Eq. \ref{vertiso_gov}.  This procedure ignores the vertical dependence of
$D=\kappa^2(z) - \sigma^2$. We show here that this has no significant
consequence for thin disks.

@@

% \section{Alternative inference of instability in the presence of weak
%   vertical shear}\label{pert_theory}
% We can also investigate the destabilizing effect of vertical shear
% without explicitly solving the full governing equation,
% Eq. \ref{iso_ode3}. We first
% consider a system with $q\equiv0$, for which the eigenfunctions are
% Hermite polynomials and the real eigenfrequencies are known. We then
% perturb this system by introducing a weak vertical shear ($|q|\ll1$)
% and linearize the governing equation. This procedure is
% \begin{align}   
%   q \to 0 + \delta q,\quad
%   W \to \He_n + \delta W,\quad
%   \hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}. 
% \end{align}
% %where for this exercise we restore $\sigma$ as the eigenfrequency.

% Linearizing the integral relation Eq. \ref{integral_relation1} in the
% thin-disk limit, and taking the
% imaginary part, we have
% \begin{align}
%   2\hat{\sigma}\imag(\delta\hat{\sigma})
%   \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
%   \He_n^2(\hat{z}) d\hat{z}
%   = \delta q \epsilon \hat{k} 
%   \int_{-\infty}^{\infty}
%   w(\zhat)\hat{z}\He_n(\hat{z})\He_n^\prime(\zhat) d\hat{z}
% \end{align}
% Recognizing $\hat{z} = \He_1(\hat{z})$ and using $\He_n^\prime = n
% \He_{n-1}$ we have
%  \begin{align}
%    2\hat{\sigma}\imag(\delta\hat{\sigma})
%    \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
%    \He_n^2(\hat{z}) d\hat{z}
%    =\delta q \epsilon \hat{k} 
%    \int_{-\infty}^{\infty}
%    w(\zhat)\He_1(\hat{z})\He_n(\hat{z})n\He_{n-1}(\hat{z})d\hat{z}. 
%  \end{align}

% Finally, we evaluate the integrals using the results
% \begin{align}
%   \int_{-\infty}^{\infty}
%   w(\xi) \He_k(\xi)\He_l(\xi) d\xi = \sqrt{2\pi}k!\delta_{kl} \quad
% % \end{align}
% \text{and} \quad
% % \begin{align}
%   \int_{-\infty}^{\infty}
%   w(\xi) \He_m(\xi)\He_k(\xi)\He_l(\xi) d\xi =
%   \frac{\sqrt{2\pi}m!k!l!}{(j-m)!(j-k)!(j-l)!}, 
% \end{align}
% where $j = (m+k+l)/2$. We then obtain
%  \begin{align}
%    2\hat{\sigma}\imag(\delta\hat{\sigma})
%   \left(1+\hat{k}^2\right) = \delta q \epsilon \hat{k} n.
%  \end{align}
% Inserting the original eigenfrequency $\hat{\sigma} = \pm
% \sqrt{n}(1+\hat{k}^2)^{-1/2}$ gives
% \begin{align}
%   \imag(\delta\hat{\sigma})= \pm \frac{1}{2}\delta q \epsilon
%   \sqrt{n} \frac{\hat{k}}{\sqrt{1+\hat{k}^2}}. 
% \end{align}
% This result agrees with Eq. \ref{simple_growth} in the limit
% $|q|\to0$. 
% %for $n=1$, since the
% %function $\He_1$ solves the governing equation exactly. 
% For $\hat{k}\gg 1$ we have
% \begin{align}
%   \imag(\delta\hat{\sigma})\simeq \pm \frac{1}{2}\delta q \epsilon
%   \sqrt{n} \sgn{\hat{k}}. 
% \end{align}

% Differentiating Eq. \ref{adia_iso1} properly gives
% \begin{align}
%   \left(D + \gamma k_x^2 c_s^2\right)\frac{d\Delta}{dz}=& D
%   \frac{d^2\delta v_z}{dz^2} + k_x^2c_s^2\left[\frac{\gamma}{\left(D + \gamma k_x^2 c_s^2\right)}\frac{dD}{dz} + \ii
%     \frac{d\ln\rho}{dz}\left(\ii +
%       \frac{q}{k_xr}\right)\right]\frac{d\delta v_z}{dz}\notag\\
%   &+ \ii k_x^2 c_s^2 \left(\ii +
%       \frac{q}{k_xr}\right)\left[\frac{d^2\ln\rho}{dz^2} -
%       \frac{1}{\left(D + \gamma
%           k_x^2c_s^2\right)}\frac{dD}{dz}\frac{d\ln\rho}{dz}\right]\delta
%     v_z. \label{adia_diso1}
% \end{align}
% The terms that were ignored in deriving Eq. \ref{adia_iso3} are those
% proportional to $dD/dz$ in Eq. \ref{adia_diso1}. Now, in the global
% disk with $\Gamma=1$ we have
% \begin{align}\label{dkappa2}
%   \frac{\p\kappa^2}{\p z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
%     r}\frac{\p\Omega^2}{\p z} = -
%   \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2} \left(2 + q +
%     \frac{z^2/r^2-2}{z^2/r^2+1}\right). 
% \end{align}
% For the local problem we may then write
% \begin{align}
%   \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z;q),
% \end{align}
% where the function $F$ corresponds to the bracket in
% Eq. \ref{dkappa2}. This function varies from $F=q$ at $z=0$ to $F\to
% 3+q$ as $|z|\to\infty$, i.e. its magnitude is of order unity. 
% % We estimate the
% % importance of these terms by noting that 
% % \begin{align}
% %   \frac{dD}{dz}\equiv \frac{d\kappa^2}{dz} \simeq \frac{d\Omega^2}{dz}
% %   = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2},
% % \end{align}
% % for thin disks. 
% Eq. \ref{adia_diso1} becomes
% \begin{align}
% \left(D + \gamma k_x^2 c_s^2\right)\frac{d\Delta}{dz}=& D
%   \frac{d^2\delta v_z}{dz^2} - k_x^2c_s^2\frac{d\ln\rho}{dz}\left[ 1 - 
%       \frac{\ii q}{k_xr}  +  \frac{\gamma q c_s^2F}{r^2\left(D +
%         \gamma k_x^2 c_s^2\right)}\right]\frac{d\delta
%     v_z}{dz}\notag\\ 
%   &- k_x^2 c_s^2 \left(1 - 
%       \frac{\ii q}{k_xr}\right)\left[\frac{d^2\ln\rho}{dz^2} + 
%       \frac{qc_s^2F}{r^2\left(D + \gamma
%           k_x^2c_s^2\right)}\left(\frac{d\ln\rho}{dz}\right)^2\right]\delta
%     v_z. \label{adia_diso2}
% \end{align}
% Since $D\sim \Omega_k^2$ in the low-frequency limit, we 
% see from Eq. \ref{adia_diso2} that the neglected terms are
% $O(\epsilon^2)$. We can make this explicit by combining
% Eq. \ref{adia_diso1}---\ref{adia_diso2} and Eq. \ref{adia_iso2}, then
% set $D\to\Omega_k^2$. In terms of non-dimensional variables, the result
% is  
% \begin{align}
%    &\delta v_z ^{\prime\prime} + \left(1 + \ii\epsilon q\hat{k} -
%     \frac{\gamma q \epsilon^2\hat{k}^2F}{1+\gamma
%       \hat{k}^2}\right)\ln\rho^{\prime}\delta v_z^\prime +
%   \left[\left(\frac{1}{\gamma} + \ii \epsilon q
%       \hat{k}\right)\ln\rho^{\prime\prime} - \hat{k}^2\left(1 -
%       \frac{\ii\epsilon
%         q}{\hat{k}}\right)\left(\frac{\gamma-1}{\gamma} +
%       \frac{q\epsilon^2F}{1+\gamma\hat{k}^2}\right)\ln\rho^{\prime
%       2}\right]\delta v_z \notag\\&=
%   -\hat{\sigma}^2\left(\frac{1}{\gamma} + \hat{k}^2\right)\delta v_z\label{adia_diso3}.
% \end{align}   
% Eq. \ref{adia_diso3} differs from Eq. \ref{adia_iso3} by terms
% proportional to $\epsilon^2$. For a thin disk, $\epsilon\ll1$, so
% neglecting these terms has no qualitative effect on the discussion in
% \S\ref{analytic_adia}.

% \section{Axisymmetric instability in vertically 
%   isothermal disks in global cylindrical geometry}\label{global_vertiso}
% Consider a locally isothermal disk in 3D such that $P=c_s^2\rho$ at
% all times, so in the perturbed state $\delta P =
% c_s^2\delta\rho$, with $c_s$ begin a prescribed function of position. 
% The axisymmetric linearized fluid equations in cylindrical
% geometry is then 
% \begin{align}
%  \ii\sigma \delta v_r + 2\Omega\delta v_\phi &= c_s^2\frac{\p}{\p r}\left(\frac{\delta\rho}{\rho}\right),\\
%  \ii\sigma \delta v_\phi &= \frac{1}{r}\delta\bm{v}\cdot\nabla j,\\
%  \ii\sigma \delta v_z &= c_s^2\frac{\p}{\p
%    z}\left(\frac{\delta\rho}{\rho}\right),\\
%  \ii\sigma\delta\rho &= \nabla\cdot\left(\rho\delta\bm{v}\right),
% \end{align}
% and we recall $j=r^2\Omega$. We can eliminate $\delta v_\phi$ from these equations to obtain
% \begin{align}
%   -\sigma^2\rho\left(|\delta v_r|^2 + |\delta v_z|^2\right) +
%   \kappa^2\rho|\delta v_r|^2 + \frac{2\Omega}{r}\frac{\p j}{\p z} \rho
%   \delta v_r^*\delta v_z =
%   c_s^2\rho\delta\bm{v}^*\cdot\nabla\left(\frac{\ii\sigma\delta\rho}{\rho}\right)
%   =c_s^2\rho\delta\bm{v}^*\cdot\nabla\left[\frac{1}{\rho}\nabla\cdot\left(\rho\delta\bm{v}\right)\right].\label{global_integral} 
% \end{align}
% At this point if we make the anelastic approximation by assuming
% $\nabla\cdot\left(\rho\delta\bm{v}\right)$ is negligible in this
% equation \citep[as argued by][]{nelson13}, 
% then the only contribution to instability comes from the third term on
% the left hand side, i.e. the explicit vertical shear term
% $\p_zj$. This is the same conclusion reached in
% \S\ref{integral_relation}.

% The term proportional to $\kappa^2$ is stabilizing. Because radial shear
% is much larger than vertical shear, this implies that significant
% instability can only be achieved with large vertical
% motions (i.e. $|\delta v_z|\gg|\delta v_r|$).

% If we integrate Eq. \ref{global_integral} over the fluid volume 
% and assume boundary terms vanish when integrating by parts, we obtain 
% \begin{align}
%   -\sigma^2\int \rho\left(|\delta v_r|^2 + |\delta v_z|^2\right) dV +
%   \int \kappa^2\rho|\delta v_r|^2 dV + \int \frac{2\Omega}{r}\frac{\p j}{\p z} \rho
%   \delta v_r^*\delta v_z dV = -\int
%   \frac{c_s^2}{\rho}\left|\nabla\cdot\left(\rho\delta\bm{v}\right)\right|^2
%   dV - \int \nabla\cdot\left(\rho\delta\bm{v}\right) \delta
%   \bm{v}^*\cdot\nabla c_s^2.
% \end{align}
% Note that the treatment so far is in fact valid for any prescribed
% $c_s(r,z)$. However, to proceed further we specialize to the case
% $c_s=c_s(r)$ for simplicity, i.e. vertically isothermal
% disks. Recalling $r\p_z\Omega^2 = -\p_z\ln{\rho}\p_r
% c_s^2$  for a vertically isothermal disk (Eq. \ref{vertical_shear}), we obtain 
% \begin{align}
%   \sigma^2\int \rho\left(|\delta v_r|^2 + |\delta v_z|^2\right) dV =
%   \int \kappa^2\rho|\delta v_r|^2 dV + \int
%   \frac{c_s^2(r)}{\rho}\left|\nabla\cdot\left(\rho\delta\bm{v}\right)\right|^2
%   dV + \int \left[ \rho \left(\nabla\cdot\delta\bm{v}\right)\delta
%     v_r^* + |\delta v_r|^2\frac{\p \rho}{\p r}\right]\frac{dc_s^2(r)}{dr}  dV.
% \end{align}
% We see that instability is only possible if
% $dc_s^2/dr\neq0$. Furthermore, since density and temperature typically
% decrease outwards in astrophysical disks, instability is associated
% with the term proportional to $\nabla\cdot\delta\bm{v}$. It is also
% clear that $\sigma^2$ will be complex in general.

@@

>>>>

\section{Applicability of the vertically global shearing box}\label{global_corr}
In \S\ref{analytical} we used the vertically global shearing box
\citep[VGSB,][]{mcnally14} to make analytic progress.   
The main difference between the linear problem in the VGSB 
and the radially-local approximation
(Eq. \ref{lin_mass}---\ref{lin_energy}) adopted in our numerical study
is the absence of  terms associated with the radial disk structure in
the former framework. These terms appear in the energy, mass and radial momentum
equations and are reproduced here for convenience,

\begin{align}
  \ii \sigma W  &=c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{\gamma}{\Gamma} \frac{d\delta
    v_z}{dz}\right) + \left[\ii k_xc_s^2(r_0,z)
  \frac{\gamma}{\Gamma} + \hat{g}_c\left.\frac{1}{\rho}\frac{\p P}{\p
      r}\right|_0\right]\delta v_x  +
\frac{1}{t_c}\left(W-\frac{Q}{\Gamma}\right),\label{gcorr_terms1}\\
\ii \sigma Q &= c_s^2(r_0,z)\left(\ii k_r + \hat{g}_c
  \left.\frac{\p\ln{\rho}}{\p r}\right|_0\right)\delta v_x + c _s^2(r_0, z)\left( \left.\frac{\p\ln\rho}{\p
      z}\right|_0 \delta v_z + \frac{d\delta
    v_z}{dz}\right),\label{gcorr_terms2} \\
\ii \sigma \delta v_x & = \ii k_x W  -
\hat{g}_c\frac{1}{c_s^2(r_0,z)}\left.\frac{1}{\rho}\frac{\p P}{\p
  r}\right|_0Q - 2\Omega(r_0,z)\delta v_y,\label{gcorr_terms3}
\end{align}
where subscript $0$ denotes evaluation at $r=r_0$.

@@

\subsection{Spurious growth of adiabatic perturbations in the VGSB}\label{analytic_adia} 
An important caveat of the VGSB is that it cannot be used in the
adiabatic limit. We explain this by setting  $\beta\to\infty$ in
Eq. \ref{vertiso_gov_nondim}, which can then be written as

\begin{align}
  0 =\dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime
  +\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) 
    -\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  \hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat 
  =&  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
&+
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat
+ \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
$\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
expected, a sub-adiabatically stratified disk is stable in the absence
of vertical shear.

@@

\subsection{Effect of global radial gradients}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/gcorr_compare} 
  \caption{Growth rate of the fundamental VSI mode as a function of
    the thermal relaxation time $\beta$. The `Analytical (VGSB)' curve
    is calculated from the dispersion relation Eq. \ref{relax_disp};
    the `Numerical (VGSB)' curve is obtained by numerically solving
    Eq. \ref{lin_mass}---\ref{lin_energy} with $\hat{g}_c=0$
    (which defines the linear VGSB equations); and the `Numerical
    (Full)' curve is obtained by numerically solving
    Eq. \ref{lin_mass}---\ref{lin_energy}  with $\hat{g}_c=1$, which
    accounts for the radial disk structure.  
    The disk parameters are $\Gamma=1.011$,
    $\gamma=1.4$ and $(p,q,\epsilon)=(-1.5,-1,0.05)$. The perturbation
    wavenumber is $\khat=30$. 
    \label{gcorr_compare}}  
\end{figure}

\section{Full governing equation for vertically isothermal disks in
  the VGSB}\label{adia_improve}

In \S\ref{approx_gov} we made the replacement $D\to\Omega_k^2$ before
eliminating variables to obtain a single equation for $\delta v_z$,
Eq. \ref{vertiso_gov}.  This procedure ignores the vertical dependence of
$D=\kappa^2(z) - \sigma^2$. We show here that this has no significant
consequence for thin disks.

@@

@@

\section{Linear problem in isothermal limit}\label{iso_discuss}  
Here we examine in more detail the special case of isothermal
perturbations ($\beta\equiv 0$) in the VGSB framework.   
%Isothermal perturbations correspond to $\beta\equiv0$ ($\chi=1$) and 
In this case Eq. \ref{ode_Q} becomes   
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$), and so $  \ii\hat{\sigma}c_s\delta v_z = W^\prime $
from Eq. \ref{lin_vz}. For isothermal perturbations it is in fact
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w}. We obtain 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \ii \epsilon q \hat{k}
    f(\zhat)\right]W^\prime + \hat{\sigma}^2\left(1+\hat{k}^2\right)W,\label{iso_ode1}
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = \epsilon^2q f(\hat{z})\Omega_k^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+\epsilon^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be considered as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$. The existence of a maximum, as remarked in
\S\ref{vshear_def}, limits the growth rate of the VSI. We explicitly
demonstrate this below.

\subsection{Necessity of vertical shear for
  instability}\label{integral_relation} 
We can establish a necessary condition for instability by multiplying
Eq. \ref{iso_ode1} by $\rho W^*$ and integrate vertically from
$\zhat=\zhat_1$ to $z=\zhat_2$. We neglect boundary 
terms when integrating by parts, by assuming $W$ or
$W^\prime$ vanishes at the boundaries, or that the boundaries are 
sufficiently far away so that the boundary terms are negligible because of the
decaying background density with increasing height. Then,
\begin{align}
  \hat{\sigma}^2\left(1+\hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|W|^2d\zhat
  =\int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\zhat 
  +\ii \epsilon q \hat{k}\int_{\zhat_1}^{\zhat_2}\rho f(\zhat) W^*W^\prime d\zhat.\label{integral_relation1}
\end{align}
It follows that for instability ($\imag\hat{\sigma}>0$), it is necessary to
have $q\neq0$ or more generally $d\Omega^2/dz\neq 0$, i.e. vertical
shear. A similar conclusion can be reached in the equivalent linear  
problem in global cylindrical disk geometry.

\subsection{Maximum growth rate}   
Here we show that the growth rate is limited by the maximum vertical
shear in the domain. % Writing $\hat{\sigma} = \hat{\omega} +
% \ii\hat{\nu}$ with real $\hat{\omega}$ and $\hat{\nu}$, 
The real and imaginary parts of 
Eq. \ref{integral_relation1} are
\begin{align}
  \left(\hat{\omega}^2-\hat{\nu}^2\right)\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
  \int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\hat{z}
  \notag
  =\real\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\zhat\right], \\
   2\hat{\omega}\hat{\nu}\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z}
  =\imag\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right],
\end{align}
where we recall $\hat{\omega} = \real \hat{\sigma}$ and
$\hat{\nu}=\imag\hat{\sigma}$. 
Adding the square of these equations give
\begin{align}
  \left[|\hat{\sigma}|^2\left(1+\hat{k}^2\right)
    \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
    \int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\right]^2
  +4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}
  =\left|\ii
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
It is clear that
\begin{align}\label{sigma_finite_domain} 
  4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime
  \right|^2d\hat{z} 
  \leq\left|
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
On the left hand side of this inequality, we apply the Cauchy-Schwarz
inequality to obtain
\begin{align}
  \left( \int_{\zhat_1}^{\zhat_2}\rho
    |W|\left|W^\prime \right|d\hat{z}\right)^2\leq
  \int_{\zhat_1}^{\zhat_2}\rho 
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}.
\end{align}
On the right hand side of Eq. \ref{sigma_finite_domain} we have
\begin{align}
  \left|\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|\leq \int_{\zhat_1}^{\zhat_2}\rho
  \left|f(\hat{z})W^*W^\prime \right|d\hat{z}
  \leq
  \mathrm{max}\left(|f|\right)\int_{\zhat_1}^{\zhat_2}\rho
  |W|\left|W^\prime \right|d\hat{z},
\end{align}
where $\mathrm{max}(|f|)$ is the maximum value of $|f|$ in
$\zhat\in[\zhat_1,\zhat_2]$. Inserting these inequalities into
Eq. \ref{sigma_finite_domain} gives
\begin{align}\label{max_growth}
  |\hat{\nu}|\leq
  \frac{\epsilon |\hat{k} q|}{2\sqrt{1+\hat{k}^2}}\mathrm{max}(|f|). 
\end{align}
Recalling that $f$ represents vertical shear (Eq. \ref{fz_shear}), it
follows that the maximum possible growth rate of unstable modes,
satisfying the above boundary conditions, is limited by the maximum
vertical shear in the domain considered.

In practice, one might consider a vertical domain of a few scale 
heights in a thin disk with $|q|=O(1)$. In this case, $|\epsilon\zhat|\ll1$ and  
$f\simeq \zhat$, so that $\mathrm{max}|f| = O(1)$, implying a
maximum growth rate $O(\epsilon \Omega_k)$, consistent with numerical
results.

% For vertical domains smaller 
% than this height, the growth rate is limited by $|f|$ at the vertical
% boundaries considered. For $|\epsilon\hat{z}|\ll1$ we have $f\simeq
% \hat{z}$, in which case we can expect growth rates to increase at most linearly
% with respect to the vertical domain size.   %no faster than linear

\subsection{Thin-disk limit}
For thin disks ($\epsilon\ll1$), in which 
$\ln\rho\simeq -\zhat^2/2$ and $f(\zhat)\simeq \zhat$, 
% by expanding the background density and vertical shear profile in powers
% of $z/r$. To lowest order we obtain 
% \begin{align}
%   &\frac{\rho}{\rho_0} \simeq 
%   \exp{\left(-\frac{\hat{z}^2}{2}\right)} \equiv w(\hat{z}),\label{thin_dens}\\
%   &\frac{d\Omega^2}{d\hat{z}} \simeq \epsilon^2q\Omega_k^2\hat{z}, \label{thin_vshear}
% \end{align} 
%With this approximation and those in \S\ref{integral_relation}, 
Eq. \ref{iso_ode1} becomes  
\begin{align}\label{iso_ode3}
  W^{\prime\prime} - \left(1 + \ii q\epsilon
    \hat{k}\right)\hat{z}W^\prime  +
  \hat{\sigma}^2\left(1+\hat{k}^2\right)W = 
  0.
\end{align}
We remark that for large $|\hat{k}|$, Eq. \ref{iso_ode3} is the same as
that derived by \cite{nelson13}, although we have taken a different
route.
